---
title: "Visualizing Plastics: Deliverable from Data"
author: "Nathan Lee, Noah Nayve, Ryan Ekdahl, Taiga Sobajima"
format: html
editor: visual
embed-resources: true
code-fold: true
---

# Data Context

### Plastic (Main Data Set)

The main data set comes from Sarah Sauve, who conducted an audit based on Break Free from Plastic (BFFP), a global organization that runs annual brand audits. In these audits, volunteers around the world collect and report plastic waste. This makes the data set observational, as it relies on self-reported data submitted by volunteers each year, specifically for 2019 and 2020 in this case. However, the exact methods used by each volunteer are unknown. Each observation represents the plastic collected from a specific parent company within a given country and year. Here, Sauve intends to utilize this data set to raise plastic pollution awareness and highlight its importance. We are further supporting Sauve with our own analysis on two research questions presented later in this deliverable.

Below is a breakdown of all the variables in this data set:

- **country** (character): country of cleanup

- **year** (double): year, either 2019 or 2020

- **parent_company** (character): source of plastic

- **empty** (double): plastics counted without a brand being recorded

- **hdpe** (double): high density polyethylene count (i.e. plastic milk containers, plastic bags, bottle caps, trash cans, oil cans, plastic lumber, toolboxes, supplement containers)

- **ldpe** (double): low density polyethylene count (i.e. plastic bags, Ziploc bags, buckets, squeeze bottles, plastic tubes, chopping boards)

- **o** (double): other counts

- **pet** (double): polyester plastic count (i.e. polyester fibers, soft drink bottles, food containers)

- **pp** (double): polypropylene count (i.e. flower pots, bumpers, car interior trim, industrial fibers, carry-out beverage cups, microwavable food containers, DVD keep cases)

- **ps** (double): polystyrene count (i.e. toys, video cassettes, ashtrays, trunks, beverage/food coolers, beer cups, wine and champagne cups, carry-out food containers, Styrofoam)

- **pvc** (double): PVC plastic count (window frames, bottles for chemicals, flooring, plumbing pipes)

- **grand_total** (double): grand total count (all types of plastic)

- **num_events** (double): number of counting events

- **volunteers** (double): number of volunteers

Before combining two CSV files, the 2019 and 2020 data, some data cleaning steps were implemented. The 2019 CSV had two columns for "pp" (Polypropylene) and "ps" (Polystyrene) counts, so the repeated columns were combined to make one overall column for each variable. The data type of the "grand_total" column from the 2020 CSV was changed from character to numeric. Lastly, some variables were renamed to match one another, ensuring no errors when combining the two CSV files.

However, this data set has important limitations and potential inconsistencies to consider. The data depends on where volunteers are located, meaning it is more likely to be collected in accessible areas, which can lead to over- or under- representation in some regions and introduce bias. Due to the nature of self-reported data, there is no way for us to verify the data. There may also be variability in how the plastic data is classified and reported to BFFP despite clear definitions.

### Country (Outside Data Set)

The supplementary data set comes from API Ninjas, a software company that publishes real-world data into formatted APIs. The Country API aggregates geographic, demographic, and economic metrics for every country in the world which can be found at: <https://api-ninjas.com/api/country>. Each observation represents a single country, where each row contains nation level statistics for that country. For example, a single row might represent the United States, containing values such as a population of approximately 331 million, a GDP of roughly \$23 trillion, and an unemployment rate around 5%. We are utilizing this API to join demographic and economic context to the primary plastics data set, supporting our research questions on how volunteer participation and plastic collection patterns vary across countries of different sizes and development levels.

The data were accessed from our code using HTTP GET requests to the /v1/country endpoint, with each country returned as a JSON object. Requests were made by country name, iterating over the unique country names in the plastics data set. Each JSON object was then joined to the plastics data set on country name, adding each country's demographic and economic metrics to its corresponding plastic collection observation.

Below is a breakdown of the key variables used from this data set:

- **name** (character): name of the country 

- **population** (integer): total population of the country

- **gdp** (double): gross domestic product in US Dollars 

- **gdp_growth** (double): GDP growth rate in % 

- **unemployment** (double): unemployment rate in % 

- **surface_area** (double): total surface area of the country in km² 

- **urban_pop_rate** (double): percentage of the population living in urban areas 

- **fertility** (double): average number of children born per woman 

- **infant_mortality** (double): number of infant deaths per 1,000 live births 

- **currency** (character): 3-letter currency code of the country 

- **threatened_species** (integer): number of threatened species

- **forested_area** (double): percentage of the country's total land area covered by forest

This data set also has important limitations to consider. API Ninjas does not include collection dates or timestamps for individual records, so we are unable to determine how recent each country's data was collected or updated. Additionally, since the exact source for each variable is not listed, it is difficult to determine the reliability of certain metrics collected for each country. There were also discrepancies in how each country name was displayed between API Ninjas and our plastics data, requiring standardization prior to merging.

# Preparing Data

### Loading in Libraries

```{r}
#| label: Loading Libraries
#| message: false

library(tidyverse)
library(glue)
library(gt)
library(plotly)
library(scales)
```

### Reading Data Sets

```{r}
#| label: Reading Plastic Data Set
#| message: false

plastics <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

# Research Questions

1.  Do countries with more volunteers tend to collect more proportions of both high and low density polyethylene plastic in 2019 and 2020?
2.  Do countries with greater forest coverage and more threatened species show a different breakdown of plastic types?

# Summaries and Visualizations

### Question 1

This custom function "plastic_volunteer_summary" finds volunteer and polyethylene plastic statistics based on an inputted polyethylene plastic type and country name. The default data set input is the main, plastic data set.

Below are the outputs generated by this function. Note, the output statistics assume a given country provided by the input and years 2019 and 2020. The letter in the () will be either "H" or "L", indicating high or low density polyethylene respectively (HDPE or LDPE):

- Ave\_()DPE (double): average polyethylene across parent companies

- Percent\_()DPE (double): percent polyethylene collected over all plastics

- Total\_()DPE (double): total polyethylene collected

- Ave_Vol (double): average number of volunteers

- Total_Vol (double): total number of volunteers

```{r}
#| label: plastic_volunteer_summary function

plastic_volunteer_summary <- function(polyethylene, country_name, data = plastics) {
  # check to see if polyethylene and country_name is in the data frame
  if (
    !(polyethylene %in% c("ldpe", "hdpe"))
    ) {
    stop(
      paste("Plastic type not found or is not polyethylene kind:", polyethylene)
      )
    }
  
  if (
    !(country_name %in% data$country)
    ) {
    stop(
      paste("Country not found:", country_name)
      )
    }
  
  # setting column names for output
  caps_dpe <- toupper(polyethylene)
  avg_col <- glue("Ave_{caps_dpe}")
  percent_col <- glue("Percent_{caps_dpe}")
  total_col <- glue("Total_Plastic")
  
  # main body finding summary stats of volunteer and plastic type
  result <- data |>
    filter(country == country_name) |>
    mutate(percent_plastic_type = .data[[polyethylene]] / grand_total) |>
    summarise(
      !!avg_col := mean(.data[[polyethylene]], na.rm = TRUE),
      !!percent_col := sum(percent_plastic_type, na.rm = TRUE) /
                        sum(!is.na(percent_plastic_type)) * 100,
      !!total_col := sum(grand_total, na.rm = TRUE),
      Ave_Vol = first(volunteers) / n(),
      Total_Vol = first(volunteers)
    ) |>
    mutate(
      across(where(is.numeric), ~round(., 2)),
      !!percent_col := paste0(.data[[percent_col]], "%")
    )
  
  return(result)
}
```

Below are high density polystyrene (HDPE) and volunteer statistics from United States over 2019 and 2020.

```{r}
#| label: hdpe for USA

plastic_volunteer_summary("hdpe", "United States of America") |>
  gt() |>
  tab_header(
    title =  md("**High Density Polystyrene Summary**"),
    subtitle = "United States of America") |>
  tab_caption("Table 1: Summary statistics for high density polystyrene plastic collected in the USA") |>
  tab_spanner(
    label = "HDPE Metrics",
    columns = c(Ave_HDPE, Percent_HDPE, Total_Plastic)
    ) |>
  tab_spanner(
    label = "Volunteer Metrics",
    columns = c(Ave_Vol, Total_Vol)
    ) |>
  cols_label(
    Ave_HDPE = "Avg HDPE",
    Percent_HDPE = "% HDPE",
    Total_Plastic = "Total Plastic",
    Ave_Vol = "Avg Volunteers",
    Total_Vol = "Total Volunteers"
  ) |>
  tab_style(
    style = cell_fill(color = "#457B9D"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "#1D3557"),
    locations = cells_column_spanners()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_spanners()
  )
```

Table 1 shows high density HDPE and volunteer statistics for the United States of America. Interestingly, the USA has both small amounts of volunteers and HDPE plastic collected at 9.2% and a mean of 0.72 HDPE plastics collected across parent companies. This may entail that the number of volunteers is correlated to the amount of HDPE plastic collected. To find whether this 9.2% is truly low, we will look at low density polystyrene (LDPE) statistics.

```{r}
#| label: ldpe for USA

plastic_volunteer_summary("ldpe", "United States of America") |>
  gt() |>
  tab_header(
    title =  md("**Low Density Polystyrene Summary**"),
    subtitle = "United States of America") |>
  tab_caption("Table 2: Summary statistics for low density polystyrene plastic collected in the USA") |>
  tab_spanner(
    label = "LDPE Metrics",
    columns = c(Ave_LDPE, Percent_LDPE, Total_Plastic)
    ) |>
  tab_spanner(
    label = "Volunteer Metrics",
    columns = c(Ave_Vol, Total_Vol)
    ) |>
  cols_label(
    Ave_LDPE = "Avg LDPE",
    Percent_LDPE = "% LDPE",
    Total_Plastic = "Total Plastic",
    Ave_Vol = "Avg Volunteers",
    Total_Vol = "Total Volunteers"
  ) |>
  tab_style(
    style = cell_fill(color = "#457B9D"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "#1D3557"),
    locations = cells_column_spanners()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_spanners()
  )
```

Even with the same number of volunteers in the USA, the proportion of LDPE collected was reported to be higher. Table 2 shows these statistics, with 14.68% of the plastic being of the LDPE type and more than 1 LDPE plastic being collected for every parent company, on average. There may be an association or more comparisons to further investigate, but we are interested in comparing a type of polystyrene across different countries, specifically looking at volunteer statistics. We decided to look into Indonesia HDPE statistics.

```{r}
#| label: hdpe for Indonesia

plastic_volunteer_summary("hdpe", "Indonesia") |>
  gt() |>
  tab_header(
    title =  md("**High Density Polystyrene Summary**"),
    subtitle = "Indonesia") |>
  tab_caption("Table 3: Summary statistics for high density polystyrene plastic collected in Indonesia") |>
  tab_spanner(
    label = "HDPE Metrics",
    columns = c(Ave_HDPE, Percent_HDPE, Total_Plastic)
    ) |>
  tab_spanner(
    label = "Volunteer Metrics",
    columns = c(Ave_Vol, Total_Vol)
    ) |>
  cols_label(
    Ave_HDPE = "Avg HDPE",
    Percent_HDPE = "% HDPE",
    Total_Plastic = "Total Plastic",
    Ave_Vol = "Avg Volunteers",
    Total_Vol = "Total Volunteers"
  ) |>
  tab_style(
    style = cell_fill(color = "#457B9D"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "#1D3557"),
    locations = cells_column_spanners()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_spanners()
  )
```

According to table 3, Indonesia has significantly more volunteers (6850 compared to USA's 740). However, the amount of HDPE plastic collected was almost half, only 4.25% of the plastic collected was of the HDPE kind. This suggest environmental and other country-specific conditions playing a major factor in different types of plastic collected, or even the distribution in plastic types. To easier see patterns and more statistics, we gathered data for the top and bottom 5 average volunteers for both high and low density.

```{r}
#| label: top and bottom 5 average volunteer hdpe and ldp stats

# top and bottom 5 countries
top5_countries <- plastics |>
  group_by(country) |>
  summarise(ave_vol = first(volunteers) / n()) |>
  slice_max(ave_vol, n = 5) |>
  pull(country)

bot5_countries <- plastics |>
  group_by(country) |>
  summarise(ave_vol = first(volunteers) / n()) |>
  slice_min(ave_vol, n = 5) |>
  pull(country)

# hdpe table
top5_hdpe <- map(top5_countries, ~plastic_volunteer_summary("hdpe", .x) |>
                   mutate(Country = .x, .before = 1)) |>
  list_rbind()
bot5_hdpe <- map(bot5_countries, ~plastic_volunteer_summary("hdpe", .x) |>
                   mutate(Country = .x, .before = 1)) |>
  list_rbind()

bind_rows(
  top5_hdpe |> mutate(Group = "Top 5", .before = 2),
  bot5_hdpe |> mutate(Group = "Bottom 5", .before = 2)
) |>
    mutate(
      across(where(is.numeric), ~ifelse(is.nan(.), 0, .)),
      across(where(is.character), ~ifelse(. == "NaN%", "0%", .))
      ) |>
  gt() |>
  tab_header(
    title =  md("**High Density Polystyrene Summary**"),
    subtitle = "Top and Bottom 5 Average Volunteer Countries") |>
  tab_caption("Table 4: Summary statistics for high density polystyrene plastic collected in the top and bottom 5 volunteer countries") |>
  tab_spanner(
    label = "Location",
    columns = c(Country, Group)
  ) |>
  tab_spanner(
    label = "HDPE Metrics",
    columns = c(Ave_HDPE, Percent_HDPE, Total_Plastic)
    ) |>
  tab_spanner(
    label = "Volunteer Metrics",
    columns = c(Ave_Vol, Total_Vol)
    ) |>
  cols_label(
    Ave_HDPE = "Avg HDPE",
    Percent_HDPE = "% HDPE",
    Total_Plastic = "Total Plastic",
    Ave_Vol = "Avg Volunteers",
    Total_Vol = "Total Volunteers"
  ) |>
  tab_style(
    style = cell_fill(color = "#457B9D"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "#1D3557"),
    locations = cells_column_spanners()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_spanners()
  )

# ldpe table
top5_ldpe <- map(top5_countries, ~plastic_volunteer_summary("ldpe", .x) |>
                   mutate(Country = .x, .before = 1)) |>
  list_rbind()
bot5_ldpe <- map(bot5_countries, ~plastic_volunteer_summary("ldpe", .x) |>
                   mutate(Country = .x, .before = 1)) |>
  list_rbind()

bind_rows(
  top5_ldpe |> mutate(Group = "Top 5", .before = 2),
  bot5_ldpe |> mutate(Group = "Bottom 5", .before = 2)
) |>
    mutate(
      across(where(is.numeric), ~ifelse(is.nan(.), 0, .)),
      across(where(is.character), ~ifelse(. == "NaN%", "0%", .))
      ) |>
  gt() |>
  tab_header(
    title =  md("**Low Density Polystyrene Summary**"),
    subtitle = "Top and Bottom 5 Average Volunteer Countries") |>
  tab_caption("Table 5: Summary statistics for low density polystyrene plastic collected in the top and bottom 5 volunteer countries") |>
  tab_spanner(
    label = "Location",
    columns = c(Country, Group)
  ) |>
  tab_spanner(
    label = "LDPE Metrics",
    columns = c(Ave_LDPE, Percent_LDPE, Total_Plastic)
    ) |>
  tab_spanner(
    label = "Volunteer Metrics",
    columns = c(Ave_Vol, Total_Vol)
    ) |>
  cols_label(
    Ave_LDPE = "Avg LDPE",
    Percent_LDPE = "% LDPE",
    Total_Plastic = "Total Plastic",
    Ave_Vol = "Avg Volunteers",
    Total_Vol = "Total Volunteers"
  ) |>
  tab_style(
    style = cell_fill(color = "#457B9D"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "#1D3557"),
    locations = cells_column_spanners()
  ) |>
  tab_style(
    style = cell_text(color = "white", weight = "bold"),
    locations = cells_column_spanners()
  )
```

There seems to be no clear association based on looking at the top and bottom 5 average volunteer countries. In table 4, Taiwan (Republic of China), Bhutan, and Benin had no HDPE plastic types reported, as opposed to Cote D'Ivoire which had 100% HDPE report rate despite also being in the top 5 in average volunteers. Based on these percentages and averages, there seems to be no clear relationship between average number of volunteers and proportion of HDPE plastic reported. The LDPE data in table 5 is similar in a sense no clear pattern is shown. However, 4 out of the top 5 average volunteer countries did have no LDPE plastic reported.

```{r}
#| label: bubble plot showing average volunteers by hdpe reported across countries
#| fig-alt: "Bubble plot showing average volunteers by the percentage high density polystyrene type reported across countries, additionally with country statistics when hovering over each circle. Note, the average volunteers are log-scaled to better fit the graph."

all_countries <- plastics |>
  distinct(country) |>
  pull(country)

bubble_data <- map(all_countries, ~plastic_volunteer_summary("hdpe", .x) |>
                     mutate(Country = .x)) |>
  list_rbind() |>
  mutate(Percent_HDPE = as.numeric(str_remove(Percent_HDPE, "%"))) |>
  filter(Percent_HDPE < 100)

bubble_plot <- ggplot(bubble_data, aes(x = Ave_Vol,
                        y = Percent_HDPE,
                        size = Total_Plastic,
                        color = Country,
                        text = paste("Country:", Country,
                                     "<br>Average Volunteers:", Ave_Vol,
                                     "<br>Percent HDPE:", Percent_HDPE, "%",
                                     "<br>Total Plastic:", comma(Total_Plastic))
                        )) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(2, 15)) +
  scale_x_log10() +
  scale_color_viridis_d() +
  guides(color = "none") +
  theme_minimal() +
  labs(
    title = "Figure 1: Average Volunteers vs. HDPE Plastic (%) by Country",
    subtitle = "Bubble size represents total plastic collected",
    x = "Average Volunteers (log scale)",
    y = "",
    size = "Total Plastic"
  )

ggplotly(bubble_plot, tooltip = "text")
```

The bubble plot in figure 1 plots the average volunteers by the percent HDPE reported across all countries in the data set. The x-axis was log-scaled to better fit the graph and show the relationship easier. There seems to be a general positive association, meaning as the average number of volunteers increase, the number of HDPE reported increases. This may be due to how common the HDPE plastic is, so naturally more people would mean more proportion of HDPE being reported. However, some countries appear twice in this plot, such as Nigeria. Due to inconsistent naming, some countries are not represented well, but the general positive trend still remains. Having more volunteers also does not necessarily mean more HDPE reports. Take Benin as an example where the average number of volunteers was the highest even though there were zero HDPE reports. These outliers may arise from the differences each country has, such as certain laws advocating for plastic reuse or rules that ban these plastic types.

### Question 2

blah blah blah
